## Plotting Cookbook

This appendix will provide `ggplot` example R code and output for of all the graphs that we might use this term. For further information, I highly recommend Kieran Healy's [Data Visualization](https://kieranhealy.org/publications/dataviz/) book and Hadley Wikham's [ggplot2](https://www.springer.com/us/book/9780387981413) book.

All the examples provided will use the standard example datasets that we have been working with throughout the term.

### Barplots

Oddly, barplots are one of the trickiest graphs in `ggplot`if you want proportions rather than absolute frequencies for each category. To make proportions, I need to specify `y=..prop..` in the aesthetics, and then use `group=1` to ensure that all categories are part of a single group and thus have proportions that sum up to one. In the code below, I am also using the `scale_y_continuous` command and the `scales::percent` command to label the tickmarks on the y-axis as percents rather than proportions. 

```{r barplot, fig.cap="A barplot", echo=TRUE}
ggplot(titanic, aes(x=pclass, y=..prop.., group=1))+
  geom_bar()+
  scale_y_continuous(labels=scales::percent)+
  labs(x="passenger class", y=NULL, 
       title="Distribution of passenger class on Titanic")+
  theme_bw()
```

The result of this code is shown in Figure \@ref(fig:barplot). Because the tickmark labels are self-explanatory I use a value of `NULL` for the y-axis label. 

### Histograms

Histograms are a snap in `ggplot`. The `binwidth` argument will allow you to easily resize bin widths without having to provide every break as base plot does. 

```{r histogram, fig.cap="A basic histogram", echo=TRUE}
ggplot(movies, aes(x=Runtime))+
  geom_histogram(binwidth = 5, col="black")+
  labs(x="movie runtime (minutes)", 
       title="Histogram of movie runtime")+
  theme_bw()
```

Figure \@ref(fig:histogram) shows the basic histogram from the code above. By default, `ggplot` will not place a border around bars, but I like to make the border a slightly different color (e.g. `col="black"`) to provide better visual definition. 

You can also calculate density instead of count, if you prefer, by adding the `y=..density..` option to the aesthetics. Density is the proportion of cases in a bin, divided by bin width. If you plot a histogram with density, you can also overlay this figure with a smoothed line approximing the distribution called a kernel density. These results are shown in Figure \@ref(fig:histogram-density). 

```{r histogram-density, fig.cap="A histogram with kernel density smoother", echo=TRUE}
ggplot(movies, aes(x=Runtime, y=..density..))+
  geom_histogram(binwidth = 5, col="black")+
  geom_density(col="red", fill="grey", alpha=0.5)+
  labs(x="movie runtime (minutes)", 
       title="Histogram of movie runtime")+
  theme_bw()
```

### Boxplots

The only tricky thing to a boxplot in `ggplot` is to remember that the aesthetic is `y` not `x`. Its not required, but setting `x=""` in the aesthetics and then setting the x label to `NULL` will also create a cleaner display with no x-axis. I also apply a `geom_boxplot` only argument here that colors outliers a different color.

```{r boxplot, fig.cap="A basic boxplot", echo=TRUE}
ggplot(movies, aes(x="", y=Runtime))+
  geom_boxplot(fill="grey", outlier.color = "red")+
  labs(y="movie runtime (minutes)", x=NULL,
       title="Boxplot of movie runtime")+
  theme_bw()
```

You can also try a `geom_violin` instead which basically gives you a mirrored kernel density plot. For a single variable, use `x=1` because `geom_violin` requires an `x` aesthetic. Below, I show how you can extend this into a comparative violin plot. 

```{r violin, fig.margin=TRUE, fig.cap="A violin plot", echo=TRUE}
ggplot(movies, aes(x=1, y=Runtime))+
  geom_violin(fill="grey")+
  scale_x_continuous(labels=NULL)+
  labs(x=NULL, y="movie runtime (minutes)", 
       title="Violin plot of movie runtime")+
  theme_bw()
```

### Comparative Barplots

Comparative barplots are definitely one of the most difficult cases to graph properly in `ggplot` because grouping causes all kinds of issues with calculating proportions correctly. The best approach that I have found is to use faceting to separate out the barplots of one variable by the categories of the other variable. The basic code is then identical to that for a single barplot above with the addition of a `facet_wrap` command to identify the second categorical variable to use for faceting.

```{r comp-barplot, fig.cap="A comparative barplot made by faceting", echo=TRUE}
ggplot(titanic, aes(x=survival, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~pclass)+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL, y=NULL, 
       title="Survival on the Titanic by class")+
  theme_bw()
```

Figure \@ref(fig:comp-barplot) shows the result. In this case, since survival and death are mirror images, we could convey the same information, but with much less ink by a simple dotplot of the percent surviving by passenger class. However, this would require some prepatory work to prepare the data in the way we want to display it. For the undergraduate class, you are not required to understand how to prepare this data. 

```{r comp-dotplot, fig.cap="A simple dotplot", echo=TRUE}
tab <- prop.table(table(titanic$survival, titanic$pclass),2)
tab <- subset(as.data.frame.table(tab), 
             Var1=="Survived", 
             select=c("Var2","Freq"))
colnames(tab) <- c("pclass","prop_surv")
ggplot(tab, aes(x=pclass, y=prop_surv))+
  geom_point()+
  scale_y_continuous(labels=scales::percent, limits=c(0,0.8))+
  labs(x=NULL, y=NULL, 
       title="Percent surviving Titanic by class")+
  coord_flip()+
  theme_bw()
```

Figure \@ref(fig:comp-dotplot) shows the resulting dotplot. If that is a little too spartan for you, you could try a lollipop graph, shown in Figure \@ref(fig:lollipop). The `geom_lollipop` is in the `ggalt` library.

```{r lollipop, fig.cap="A lollipop plot", echo=TRUE}
ggplot(tab, aes(x=pclass, y=prop_surv))+
  geom_lollipop()+
  scale_y_continuous(labels=scales::percent, limits=c(0,0.7))+
  labs(x=NULL, y=NULL, 
       title="Percent surviving Titanic by class")+
  coord_flip()+
  theme_bw()
```

### Comparative Boxplots

To construct comparative boxplots, we just need to add an `x` to the aesthetic of a boxplot. 

```{r comp-boxplot, fig.cap="A basic comparative boxplot", echo=TRUE}
ggplot(movies, aes(x=Genre, y=Runtime))+
  geom_boxplot(fill="grey", outlier.color = "red")+
  labs(y="movie runtime (minutes)", 
       title="Boxplot of movie runtime by genre")+
  theme_bw()
```

The mediocre results are shown in Figure \@ref(fig:comp-boxplot). The labels on the x-axis often run into one another, so `coord_flip` is a good option to avoid that. In addition, I can also use the `varwidth` argument in `geom_boxplot` to scale the width of boxplots relative to the number of observations. 

```{r comp-boxplot2, fig.cap="Flipping coordinates for better display", echo=TRUE}
ggplot(movies, aes(x=Genre, y=Runtime))+
  geom_boxplot(fill="grey", outlier.color = "red", 
               varwidth=TRUE)+
  labs(y="movie runtime (minutes)", 
       title="Boxplot of movie runtime by genre")+
  coord_flip()+
  theme_bw()
```

My improved comparative boxplot is shown in Figure \@ref(fig:comp-boxplot2). However, Its also often useful to re-order the categories by the mean or median of the quantitative variable. We can do this with the `reorder` command: 

```{r comp-boxplot3, fig.cap="Re-order categories by median", echo=TRUE}
ggplot(movies, aes(x=reorder(Genre, Runtime, median), y=Runtime))+
  geom_boxplot(fill="grey", outlier.color = "red", 
               varwidth=TRUE)+
  labs(y="movie runtime (minutes)", x=NULL, 
       title="Boxplot of movie runtime by genre")+
  coord_flip()+
  theme_bw()
```

The results are shown in Figure \@ref(fig:comp-boxplot3). In this case, we used the `reorder` command to reorder the categories of `Genre` by the median value of `Runtime`. 

We could also try a comparative violin plot as shown in Figure \@ref(fig:comp-violin).

```{r comp-violin, fig.margin=TRUE, fig.cap="A comparative violin plot", echo=TRUE}
ggplot(movies, aes(x=reorder(Genre, Runtime, median), y=Runtime))+
  geom_violin(fill="grey")+
  labs(y="movie runtime (minutes)", x=NULL, 
       title="Boxplot of movie runtime by genre")+
  coord_flip()+
  theme_bw()
```

### Scatterplots

The `geom_point` command gives us the the geom we want for scatterplots. We can also use `geom_smooth` to graph a variety of lines to our data. Setting an `alpha` value can help with overplotting as well by providing sem-transparency.

```{r scatter, fig.cap="Basic Scatterplot", echo=TRUE}
ggplot(politics, aes(x=age, y=income))+
  geom_point(alpha=0.5)+
  scale_y_continuous(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016")+
  theme_bw()
```

The results are shown in Figure \@ref(fig:scatter). As more points are plotted in the same position, the overall plot darkens. However, there is still a lot of overplotting because age is only recorded in whole numbers.  In cases like this, we can use `geom_jitter` rather than `geom_point` to randomly perturb each value pair a little bit, as shown in Figure 14.


```{r scatter-jitter, fig.cap="Jittering points to deal with overplotting", echo=TRUE}
ggplot(politics, aes(x=age, y=income))+
  geom_jitter(alpha=0.5)+
  scale_y_continuous(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016")+
  theme_bw()
```

It is often useful to apply some model to the data to draw a best-fitting line through the points. The `geom_smooth` function will do this for us and we can use the `method` argument to specify what kind of model we want to fit. You an apply non-parametric smoothers like LOESS as well as an OLS regression line. Lets first try a LOESS smoother. 

```{r scatter-loess, fig.cap="Plotting loess smoother to data", echo=TRUE}
ggplot(politics, aes(x=age, y=income))+
  geom_jitter(alpha=0.5)+
  geom_smooth(method="loess")+
  scale_y_continuous(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016")+
  theme_bw()
```

Figure \@ref(fig:scatter-loess) shows us the smoothed line. It looks like an inverted U-shaped distribution. Family income is also highly right skewed. Lets try `scale_y_log10` to put income on a log-scale (advanced).


```{r scatter-log, fig.cap="Changing scale of y-axis to logarithmic", echo=TRUE}
ggplot(politics, aes(x=age, y=income))+
  geom_jitter(alpha=0.5)+
  geom_smooth(method="loess")+
  scale_y_log10(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016")+
  theme_bw()
```

Figure \@ref(fig:scatter-log) shows the scatterplot with an exponential scale for the y-axis. Note how the values go from \$1K to \$10K to \$100K at even intervals. 

Now lets try switch the smoothing to a linear model, by simply changing the method in `geom_smooth`:

```{r scatter-lm, fig.cap="Fitting an OLS regression line", echo=TRUE}
ggplot(politics, aes(x=age, y=income))+
  geom_jitter(alpha=0.5)+
  geom_smooth(method="lm")+
  scale_y_log10(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016")+
  theme_bw()
```

Figure \@ref(fig:scatter-lm) shows the OLS line predicting income. Its pretty flat which to some extent hides the underlying curvilinear relationship. In practice, I would probably want to apply a quadratic term on age to any regression model. 

Finally, we can add add in an aesthetic for a third categorical variable by applying color to the plots. In this case, I will color the plots by highest degree received.

```{r scatter-color, fig.cap="Adding color aesthetic", echo=TRUE}
ggplot(politics, aes(x=age, y=income))+
  geom_jitter(alpha=0.5, aes(color=educ))+
  geom_smooth(method="lm")+
  scale_color_brewer(palette = "YlOrRd")+
  scale_y_log10(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016",
       color="Highest degree")+
  theme_bw()
```

Figure \@ref(fig:scatter-color) shows the result. Because I defined the color aesthetic in `geom_jitter` rather than the base `ggplot` command, I still only get one smoothed OLS regression line for the whole dataset. If I wanted separate lines by educational level, I could add the color aesthetic to `ggplot` instead.

```{r scatter-color2, fig.cap="Separate OLS lines by color aesthetic", echo=TRUE}
ggplot(politics, aes(x=age, y=income, color=educ))+
  geom_jitter(alpha=0.5)+
  geom_smooth(method="lm")+
  scale_color_brewer(palette = "YlOrRd")+
  scale_y_log10(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016",
       color="Highest degree")+
  theme_bw()
```

Figure \@ref(fig:scatter-color2) shows the same plot but now with separate lines for each educational level. We are basically plotting interaction terms here between age and educational level. 

Instead of coloration, I could have used faceting to try to separate out differences by education. Figure \@ref(fig:scatter-facet) shows the results using this approach. For this graph, I use two separate `geom_smooth` commands to get the LOESS and linear fit for each panel.  

```{r scatter-facet, fig.cap="facet by education instead of color", echo=TRUE}
ggplot(politics, aes(x=age, y=income))+
  geom_jitter(alpha=0.3)+
  geom_smooth(method="lm")+
  geom_smooth(method="loess", color="red")+
  facet_wrap(~educ)+
  scale_y_log10(labels=scales::dollar)+
  labs(x="age",
       y="family income (in thousands USD)",
       title="Scatterplot between age and family income", 
       caption="ANES 2016")+
  theme_bw()
```

If you have a small number of cases, it might make sense to label your values. You can label values with `geom_text` but you can often run into problems with labels being truncated or overlapping.

```{r scatter-label, fig.cap="labeling points can look really bad", echo=TRUE}
ggplot(crimes, aes(x=Unemployment, y=Property))+
  geom_point(alpha=0.7)+
  geom_smooth(method="lm", se=FALSE)+
  geom_text(aes(label=State), size=2)+
  labs(x="Unemployment rate", 
       y="Property crimes per 100,000 population",
       title="Scatterplot of unemployment and property crime")+
  theme_bw()
```

Figure \@ref(fig:scatter-label) shows labeling with the default `geom_text` approach. You can fiddle with a variety of arguments here that let you offset the labels but it rarely produces clean labels. The library `ggrepel` has a function `geom_text_repel` that will work harder to place labels in a readable position:

```{r scatter-repel, fig.cap="repel works better for labeling points, but still can get too busy", echo=TRUE}
ggplot(crimes, aes(x=Unemployment, y=Property))+
  geom_point(alpha=0.7)+
  geom_smooth(method="lm", se=FALSE)+
  geom_text_repel(aes(label=State), size=2)+
  labs(x="Unemployment rate", y="Property crimes per 100,000 population",
       title="Scatterplot of unemployment rate and property crime rate")+
  theme_bw()
```