--- 
title: "Statistical Analysis in Sociology"
author: "Aaron Gullickson"
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
output: 
  bookdown::pdf_book:
    number_section: false
  bookdown::gitbook:
    split_by: "section+number"
    number_section: false
site: bookdown::bookdown_site
documentclass: book
#bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
#github-repo: rstudio/bookdown-demo
description: "Textbook for Statistical Analysis for University of Oregon Sociology"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, message = FALSE, error = FALSE, fig.align="center")
```


# Preface {-}

This online textbook combines all of the information taught in my undergraduate statistics course as well as both terms of my introductory graduate statistics course. It is very much a work in progress at the moment. I am adding new chapters and transitioning most of my figures to `ggplot` so you will likely notice many quirks. 

&copy; Aaron Gullickson, 2017

This material is licensed under the Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International Public License. Users are free to remix tweak, and build upon this work for non-commercial purposes, although new work must acknowledge the original author and use the same license. Full license is available [here](https://creativecommons.org/licenses/by-nc-sa/4.0/legalcode).

```{r load-stuff, include=FALSE}
#load libraries
library(here)
source(here("check_packages.R"))
source(here("example_datasets","load_all_data.R"))
```
 
```{r check_html, include=FALSE}
is_html <- knitr::opts_knit$get("rmarkdown.pandoc.to")=="html"
```
