--- 
title: "Statistical Analysis in Sociology"
author: "Aaron Gullickson"
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
output: 
  bookdown::pdf_book:
    number_section: false
  bookdown::gitbook:
    split_by: "section+number"
    number_section: false
site: bookdown::bookdown_site
lot: yes
lof: yes
documentclass: book
#bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
#github-repo: rstudio/bookdown-demo
description: "Textbook for Statistical Analysis for University of Oregon Sociology"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, message = FALSE, error = FALSE, fig.align="center")
```


# Preface {-}

This online textbook combines all of the information taught in my undergraduate statistics course as well as both terms of my introductory graduate statistics course. The undergraduate course consists of the first five chapters/modules up to and including **Building Models**. I have put a lot of work into this book, but it is still very much a work in progress, so you may notice typos and other errors on occasion.

This textbook is designed to be used with the [R statistical software program](https://www.r-project.org/). If you are taking this course from me, then you will be running *R* through the [RStudio Cloud](https://rstudio.cloud). Various snippets of R code are interspersed throughout the book in order to show you how to do things. You will also find useful information in the appendices of this book. 

&copy; Aaron Gullickson, 2017

This material is licensed under the Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International Public License. Users are free to remix tweak, and build upon this work for non-commercial purposes, although new work must acknowledge the original author and use the same license. Full license is available [here](https://creativecommons.org/licenses/by-nc-sa/4.0/legalcode).

```{r load-stuff, include=FALSE}
#load libraries
library(here)
source(here("check_packages.R"))
source(here("example_datasets","load_all_data.R"))
```
 
```{r check_html, include=FALSE}
is_html <- knitr::opts_knit$get("rmarkdown.pandoc.to")=="html"
```
