---
title: "Politics Dataset"
author: |
 | Sociology 312, University of Oregon
 | Prof. Gullickson
output: 
  pdf_document: 
    fig_height: 3.5
    fig_width: 5
    number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stargazer)
load("politics.RData")
```

#Overview

This data comes from the 2012 [American National Election Study](http://www.electionstudies.org) (ANES). The ANES is a survey of the American electorate that is conducted every two years. The study collects information on a variety of political attitudes and voting behaviors. For our purposes, we are going to primarily look at respondent's attitudes on three issues: (1) immigration, (2) gay marriage, and (3) global warming.

Unlike our other datasets, some of the variables originally include significant number of missing values. In most cases, this is because the respondent refused to answer the question or didn't understand the question enough to feel comfortable answering the question. To simplify the analysis here, I have used some advanced techniques to impute valid values for all of the missing valuese. 

Table 1 provides summary statistics for all the quantitative variables in the politics data. 

```{r results='asis', echo=FALSE}
stargazer(politics, summary.stat=c("n","mean","sd","min","p25","median","p75","max"), header=FALSE, 
          digit.separate = 0,
          type="latex", title="Summary statistics for all quantitative variables in the politics dataset")
```

#Variable Descriptions

##immigcitizen

Respondents were asked whether they would support a proposal to grant citizenship to some (not necessarily all) undocumented immigrants. Respondents could either favor, oppose, or do neither. There are 10 missing values for this response. 

```{r echo=FALSE, fig.align='center'}
par(mar=c(3,7,1,1))
barplot(table(politics$immigcitizen), col=c("darkgreen"), las=2, cex.main=0.8, cex.names=0.75,
        main="Support for citizenship for undocument migrants", ylab="", xlab="frequency",
        horiz=TRUE,
        names.arg=c("Oppose","Neither/\nDon't Know","Favor"))
```

##gaymarriage

Respondents were asked for their position on gay marriage and were given the choices of "no legal recognition", "civil union (but no marriage)", "support gay marriage." There were 80 missing values. 

```{r echo=FALSE, fig.align='center'}
par(mar=c(3,7,1,1))
barplot(table(politics$gaymarriage), col=c("orange"), las=2, cex.main=0.8, cex.names=0.75,
        main="Support for gay marriage", ylab="", xlab="frequency", horiz=TRUE,
        names.arg=c("No legal\nrecognition","Civil union","Support gay\nmarriage"))
```

##globalwarm

I constructed this variable from two separate questions. The first question asks whether respondents think that global warming has been happening with the options being that it "probably has" or "probably has not." The second question asks whether respondents thought that global warming was caused by human activity (either entirely or partially). I combine these into a single dichotomous variable where individuals either think the earth is warming from human activity or that it is not warming from human activity, where the latter category includes people who think it isn't warming at all and people who think it is warming but not because of human activity. There are 142 missing values for this variable.

```{r echo=FALSE, fig.align='center'}
par(mar=c(3,7,1,1))
barplot(table(politics$globalwarm), col=c("red"), las=2, cex.main=0.8, cex.names=0.75,
        main="Belief about global warming", ylab="", xlab="frequency", horiz=TRUE,
        names.arg=c("Not warming from\nhuman activity","Warming from\nhuman activity"))
```

##party

The political party with which the respondent identifies. The original data include a category for "no preference" that I have collapsed with the category of "independent."

```{r echo=FALSE, fig.align='center'}
barplot(summary(politics$party), col=c("blue","red","purple","grey"), las=2, cex.main=0.8, cex.names=0.75,
        main="Party affiliation", xlab="", ylab="frequency")
```

#relig

The respondent's religion. To keep things simpler here, I have combined the category of evangelicial and "undifferentiated" protestant. The latter category includes protestants who are so-called "non-denominational." I have also collapsed black protestant denomination with mainline protestants. Thereare 86 missing values for this variable.

```{r echo=FALSE, fig.align='center'}
par(mar=c(3,7,1,1))
barplot(table(politics$relig), col=c("yellow"), las=2, cex.main=0.8, cex.names=0.75,
        main="Religious Affiliation", ylab="", xlab="frequency", horiz=TRUE,
        names.arg=c("Mainline Protestant","Evangelical/\nUndifferentiated\nProtestant",
                    "Roman Catholic", "Jewish","Other Religion","Not Religious"))
```

##age

The age of the respondent. The ANES is focused on the voting age population. The youngest interviewee was 17 years old. Presumably this person was interviewed because they would turn 18 by the time of the actual election. There are 60 missing values for this variable.

```{r echo=FALSE, fig.align='center'}
hist(politics$age, breaks=c(seq(from=15, to=90, by=5)), col="skyblue", las=1, cex.main=0.8,
     main="Histogram of age", xlab="age", ylab="frequency")
```

##race 

The race of the respondent. This is always a tricky category to code. The ANES uses a relatively simple and crude set of six categories and does not allow a respondent to check multiple races. There are 24 missing values on this variable.

```{r echo=FALSE, fig.align='center'}
par(mar=c(3,7,1,1))
barplot(table(politics$race), col=c("grey"), las=2, cex.main=0.8, cex.names=0.75,
        main="Race", ylab="", xlab="frequency", horiz=TRUE,
        names.arg=c("White","Black","Hispanic","Asian/\nPacific Islander", "American\nIndian",
                    "Other"))
```


##educ

The respondent's level of education. This is recorded as an ordinal variable. The "Some college" response indicates individuals who have attended college (including 2-year programs) but have not earned a BA. There are 50 missing values. 

```{r echo=FALSE, fig.align='center'}
par(mar=c(3,7,1,1))
barplot(table(politics$educ), col=c("gold"), las=2, cex.main=0.8, cex.names=0.75,
        main="Educational Attainment", ylab="", xlab="frequency", horiz=TRUE,
        names.arg=c("Less than\nhigh school","High school\ngraduate","Some college",
                    "BA degree","Graduate\ndegree"))
```


##income

The family income of the respondentin 1000s of dollars. Respondents did not give actual dollar amounts here but rather indicated which bracket of income (e.g. $20,000-30,000) they fell within. I am treating this like a continuous variable by assigning each respondent the midpoint of the bracket as their actual income (e.g. $20,000-30,000 becomes $25,000). The final category of $250,000 or more is treated as $250,000. There are 520 missing values for income. 

```{r echo=FALSE, fig.align='center'}
hist(politics$income, breaks=c(seq(from=0, to=260, by=10)), col="skyblue", las=1, cex.main=0.8,
     main="Histogram of age", xlab="income", ylab="frequency")
```


##workstatus

The work status of the respondents. Respondents could either be working, unemployed, or out of the labor force. The last category refers to people who are not employed and not currently looking for work, whereas unemployed indicates a person who is not employed an is currently looking for work. There are 13 missing values. 


```{r echo=FALSE, fig.align='center'}
par(mar=c(3,7,1,1))
barplot(table(politics$workstatus), col=c("grey"), las=2, cex.main=0.8, cex.names=0.75,
        main="Work Force Status", ylab="", xlab="frequency", horiz=TRUE,
        names.arg=c("Working","Unemployed","Out of\nLabor Force"))
```